---
title: 'Assignment 4: Data Wrangling'
author: "Hanna Bliska"
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
#install.packages('knitr')
library(formatR)
library(knitr)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=45), tidy=TRUE)
knitr::opts_chunk$set(root.dir="~/Desktop/EDA-Fall2022")
#used commands with format R and knitr to set format and working directory
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions
1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct7th @ 5:00pm.

## Set up your session 
1. Check your working directory, load the `tidyverse` and `lubridate` packages, and upload all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in a factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Explore the dimensions, column names, and structure of the datasets. 

```{r message=FALSE}
#1
getwd()
#install.packages("tidyverse")
library(tidyverse)
#install.packages("lubridate")
library(lubridate)
EPAair.O3.2018 <- read.csv("./Data/Raw/EPAair_O3_NC2018_raw.csv", stringsAsFactors = TRUE)
EPAair.O3.2019 <- read.csv("./Data/Raw/EPAair_O3_NC2019_raw.csv", stringsAsFactors = TRUE)
EPAair.PM25.2018 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv", stringsAsFactors = TRUE)
EPAair.PM25.2019 <- read.csv("./Data/Raw/EPAair_PM25_NC2019_raw.csv", stringsAsFactors = TRUE)

#2
#Exploring EPAair.O3.2018
dim(EPAair.O3.2018)
colnames(EPAair.O3.2018)
str(EPAair.O3.2018)

#Exploring EPAair.O3.2019
dim(EPAair.O3.2019)
colnames(EPAair.O3.2019)
str(EPAair.O3.2019)

#Exploring EPAair.PM25.2018
dim(EPAair.PM25.2018)
colnames(EPAair.PM25.2018)
str(EPAair.PM25.2018)

#Exploring EPAair.PM25.2019
dim(EPAair.PM25.2019)
colnames(EPAair.PM25.2019)
str(EPAair.PM25.2019)
```

## Wrangle individual datasets to create processed files.
3. Change date to date
4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r}
#3
EPAair.O3.2018$Date <- as.Date(EPAair.O3.2018$Date, format = "%m/%d/%Y")
class(EPAair.O3.2018$Date)
EPAair.O3.2019$Date <- as.Date(EPAair.O3.2019$Date, format = "%m/%d/%Y")
class(EPAair.O3.2019$Date)
EPAair.PM25.2018$Date <- as.Date(EPAair.PM25.2018$Date, format = "%m/%d/%Y")
class(EPAair.PM25.2018$Date)
EPAair.PM25.2019$Date <- as.Date(EPAair.PM25.2019$Date, format = "%m/%d/%Y")
class(EPAair.PM25.2019$Date)

#used as.Date function to reformat the date column from a factor to a date.

#4
AQI.EPAair.O3.2018 <- select(EPAair.O3.2018, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)

AQI.EPAair.O3.2019 <- select(EPAair.O3.2019, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)

AQI.EPAair.PM25.2018 <- select(EPAair.PM25.2018, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)

AQI.EPAair.PM25.2019 <- select(EPAair.PM25.2019, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)

#used select to create four data frames with columns of interest.

#5
AQI.EPAair.PM25.2018$AQS_PARAMETER_DESC <- "PM2.5"
AQI.EPAair.PM25.2019$AQS_PARAMETER_DESC <- "PM2.5"

#modified column AQS_PARAMETER_DESC to have PM2.5 in each row in the two PM2.5 data frames.

#6
write.csv(AQI.EPAair.O3.2018, row.names = FALSE, file= "./Data/Processed/EPAair_O3_NC2018_processed.csv")
write.csv(AQI.EPAair.O3.2019, row.names=FALSE, file=
"./Data/Processed/EPAair_O3_NC2019_processed.csv")
write.csv(AQI.EPAair.PM25.2018,row.names=FALSE, file= "./Data/Processed/EPAair_PM25_NC2018_processed.csv")
write.csv(AQI.EPAair.PM25.2019,row.names=FALSE, file=
"./Data/Processed/EPAair_PM25_NC2019_processed.csv")

#write.csv allowed me to save processed data sets. 

```


## Combine datasets
7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code. 
8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

* Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels)
* Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site, aqs parameter, and county. Take the mean of the AQI value, latitude, and longitude.
* Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
* Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.
10. Call up the dimensions of your new tidy dataset.
11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1819_Processed.csv"

```{r}
#7
EPAair_03_PM25_NC1819_Join <- rbind(AQI.EPAair.O3.2018, AQI.EPAair.O3.2019, AQI.EPAair.PM25.2018, AQI.EPAair.PM25.2019)

#used rbind function to join the O3 and PM2.5 data frames, which share identical column names. 

#8
EPAair_03_PM25_NC1819_Processed <-
  EPAair_03_PM25_NC1819_Join %>%
  filter(Site.Name == "Linville Falls" | Site.Name == "Durham Armory" | Site.Name == "Leggett"| Site.Name == "Hattie Avenue" | Site.Name == "Clemmons Middle" | Site.Name == "Mendenhall School" | Site.Name == "Frying Pan Mountain" | Site.Name == "West Johnston Co." | Site.Name == "Garinger High School" | Site.Name == "Castle Hayne" | Site.Name == "Pitt Agri. Center" | Site.Name == "Bryson City" | Site.Name == "Millbrook School") %>% 
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  summarise(meanAQI = mean(DAILY_AQI_VALUE), 
            meanlat = mean(SITE_LATITUDE),
            meanlong = mean(SITE_LONGITUDE)) %>%
  mutate(month = month(Date)) %>%
  mutate(year = year(Date))
dim(EPAair_03_PM25_NC1819_Processed)

#for this pipe, used filter to include the sites that all four data frames have in common. Then, used group_by to combine rows that had the same Date, Site.Name, AQS_PARAMETER_DESC, and COUNTY into one row and used summarise to produce the mean of the DAILY_AQI_VALUE, SITE_LATITUDE, and SITE_LONGITUDE in each unique row.
#checked dimensions

#9
EPAair_03_PM25_NC1819_Processed_Spread <- pivot_wider(EPAair_03_PM25_NC1819_Processed, names_from = AQS_PARAMETER_DESC, values_from = meanAQI) 

#used pivot_wider to spread AQI values into two columns, one for ozone and one for PM2.5. Took names from AQS_PARAMETER_DESC and values from meanAQI.  

#10
dim(EPAair_03_PM25_NC1819_Processed_Spread) #checked dimensions

#11
write.csv(EPAair_03_PM25_NC1819_Processed_Spread, row.names=FALSE, file="./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv") 
#saved processed file

```


## Generate summary tables
12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where ozone and PM2.5 are not available (use the function `drop_na` in your pipe). 

13. Call up the dimensions of the summary dataset.

```{r}
#12a
Summary.EPAair_03_PM25_NC1819 <- 
  EPAair_03_PM25_NC1819_Processed_Spread %>%
  group_by(Site.Name, month, year) %>%
  summarise(mean.AQI.ozone = mean(Ozone),
            mean.AQI.PM2.5 = mean(PM2.5))

#generated summary table using pipe. Used group_by to combine rows with the same Site.Name, month, and year and summarise to produce the mean ozone and PM2.5 values for each unique row. 

#12b
Up.Summary.EPAair_03_PM25_NC1819 <-
  Summary.EPAair_03_PM25_NC1819 %>%
  drop_na(mean.AQI.ozone, mean.AQI.PM2.5)

#used drop_na to remove NAs in mean.AQI.ozone and mean.AQI.PM2.5 columns.
          
#13
dim(Up.Summary.EPAair_03_PM25_NC1819) #checked the dimensions
  
```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: We used the function `drop_na` because we wanted to drop rows containing missing values in specific columns that we provided (in this case, O3 and PM2.5). If we were to use `na.omit` on our data frame, we would remove all NAs. In this case, using `na.omit` would yield the same data frame as `drop_na` because the only columns with NAs in our data frame were O3 and PM2.5, but if that were not the case and NAs were present in other columns, using `drop_na` would be better practice because it would allow us to only remove those NAs in specified columns. I also looked online and some articles recommended using `drop.na` if working with tidyverse, which we are doing in class.
